---
title: "Cohort-Based Enrollment and Revenue Modeling Pipeline"
author: "John Schofield"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: '2'
  pdf_document:
    toc: true
    toc_depth: '2'
  html_document:
    theme:
      version: 4
      bootswatch: darkly
    toc: true
    toc_depth: 2
    self_contained: false
---

```{r setup, include=FALSE}
# if you’re on TinyTeX, make sure bookmark.sty is installed (only needed for PDF)
if (requireNamespace("tinytex", quietly = TRUE) &&
    tinytex::is_tinytex()) {
  tinytex::tlmgr_install("bookmark")
}

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  error   = TRUE,
  tidy    = FALSE
)
```

This report processes and analyzes net tuition revenue (NTR), enrollment patterns, and retention cohorts for Berklee’s BCM and BCB programs. It integrates historical and current student data from multiple internal sources, cleans and transforms them, and produces cohort-level summaries to support strategic enrollment management and financial forecasting. Office of Student Accounts (OSA) provides the historical census snapshots for NTR.

Key tasks include: • Importing and cleaning datasets from Tableau Prep outputs and archived enrollment exports • Calculating term sequences and cohort codes using custom functions for comparing academic terms • Filtering and matching student-level records to reference mappings for accurate program, degree, and enrollment classification • Aggregating NTR components (tuition, institutional aid, comprehensive fees) by cohort, program, and term • Exporting results to Google Sheets and CSV for stakeholder access and downstream use

Additionally, the script handles alternate matching logic for records that do not conform to standard mappings, prepares clean datasets for retention modeling, and supports fiscal year forecasting by generating low/mid/high enrollment and revenue scenarios.

```{r load-libraries}
library(readr)
library(dplyr)
library(stringr)
library(tidyr)
library(janitor)
library(data.table)
library(lubridate)
library(readxl)
library(googlesheets4)
library(ggplot2)
library(gt)
```

```{r term-calcs}
term_code_to_num <- function(term) {
  term_year <- as.integer(substr(term, 1, 4))
  term_sem <- substr(term, 5, 6)
  if (is.na(term_year)) return(-1)
  to_add <- switch(term_sem, "SP" = -7, "FA" = 2, "SU" = 1, return(-1))
  return(term_year * 10 + to_add)
}

get_next_term <- function(term_num, include_summer = TRUE) {
  remainder <- term_num %% 10
  if (remainder == 3) return(term_num + ifelse(include_summer, 8, 9))
  return(term_num + 1)
}

term_int_to_str <- function(term) {
  year <- floor(term / 10)
  sem <- switch(term %% 10, "1" = "SU", "2" = "FA", "3" = {year <- year + 1; "SP"}, stop("Invalid term"))
  return(paste0(year, sem))
}

get_num_terms_between <- function(term_1, term_2, include_summer = TRUE) {
  dist <- abs(term_1 - term_2)
  tot <- if (include_summer) (dist %/% 10) * 3 + case_when(dist %% 10 == 1 ~ 1, dist %% 10 == 2 ~ 2, dist %% 10 == 8 ~ 1, dist %% 10 == 9 ~ 2, TRUE ~ 0)
  else (dist %/% 10) * 2 + case_when(dist %% 10 == 1 ~ 1, dist %% 10 == 9 ~ 1, TRUE ~ 0)
  return(tot)
}

get_terms_ahead <- function(term, terms_ahead, include_summer = TRUE) {
  years_ahead <- if (include_summer) terms_ahead %/% 3 else terms_ahead %/% 2
  for (i in seq_len(terms_ahead %% ifelse(include_summer, 3, 2))) term <- get_next_term(term)
  return(term + years_ahead * 10)
}

# Safe table override
table <- function(...) base::table(...)
```

```{r  bcm_net_revenue}

# Tableau Prep Flow:
#  Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcm_osa/merged_bcm_osa/bcm_merged_nr.tfl

bcm_merged_nr <- read_csv(
  "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcm_osa/merged_bcm_osa/bcm_merged_nr.csv"
)

# BCM NTR Logic / Code ConCat
bcm_xreference <- read_csv(
  "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcm_osa/bcm_xreference.csv"
)

# Archive Distributions
cohorts_term <- read_csv(
  "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcm_osa/cohorts_202503_trimmed.csv"
)

# Tableau Prep Flow
#  Enrollment_Marketing_Analytics/Student Financial Services Analytics/Affordability/Tableau Prep Flows/Enrollment_Archive_Flow.tfl
enrollment_archive <- read_csv(
  "/Volumes/Enrollment_Marketing_Analytics/Student Financial Services Analytics/Affordability/Exports/Enrollment_Archive.csv"
)


# Preserve Original Variable (bcm_merged_nr)
n_nr_data <- bcm_merged_nr
# Clean Headers and Remove Non-Numeric Variables
nr_1 <- n_nr_data %>%
  distinct() %>%
  clean_names() %>%
  # Select columns where the names do not contain numbers
  select(matches("^[^0-9]*$|.*_rate$")) %>%
  # Create Unique Term Key
  rename(term_key = bcm_nr_term_key)
str(nr_1)
# Handle Non-Degree's and Misc Student Types
nr_2 <- nr_1 %>%
  filter(!is.na(student_type) | !is.na(program_active)) %>%
  filter(
    program_active != "UG.POST" &
      program_active != "GRAD.UG" &
      program_active != "SNM.MISC" &
      program_active != "SNM.FAST" & enroll_source != "INA"
  ) %>%
  distinct()
dim(nr_2)
# Ensure the number of term_keys matches the number of rows
{
  length(unique(nr_2$term_key))
  length(nr_2$term_key)
  length(nr_2$term_key) - length(unique(nr_2$term_key))
}
nr_3 <- nr_2
str(nr_3)


# Join NTR dataset with historical cohorts dataset
nr_4 <- nr_3 %>%
  left_join(cohorts_term, by = "id_stu")

# Take only term_key and matric source from the archives
arch_1 <- enrollment_archive %>%
  select(Term_Key, `BCM Matriculation Source`) %>%
  clean_names() %>%
  distinct()

# Join archive to NTR and Cohort by term_key
nr_4_archive <- nr_4 %>%
  left_join(arch_1, by = "term_key")


# Clean and Calculate Matriculation Source Summary Field
cols_nr <- nr_4_archive %>%
  mutate(
    enroll_source_tmpl = case_when(
      term == cohort ~ "ENT",
      bcm_matriculation_source == "ENT" ~ "ENT",
      bcm_matriculation_source == "CON" ~ "CON/RET",
      bcm_matriculation_source == "RET" ~ "CON/RET",
      enroll_source == "ENT" ~ "ENT",
      enroll_source == "CON" ~ "CON/RET",
      enroll_source == "RET" ~ "CON/RET"
    )
  )
table(cols_nr$enroll_source_tmpl)

# ConCat Program/Degree/Enrollment - Source Student Program + Student Type + enroll_source_tmpl
cols_nr_2 <- cols_nr %>%
  mutate(
    pgm_dgr_enroll_tmpl = paste0(student_program, student_type, enroll_source_tmpl),
    alt_pgm_dgr_enroll_tmpl = paste0(pgm_dgr_enroll_tmpl, program_active)
  )

# 1 of 2 Join - left_join with xref table
joined <- cols_nr_2 %>%
  left_join(bcm_xreference, by = "pgm_dgr_enroll_tmpl")

# Remove Unmatched
matched <- joined %>%
  filter(!is.na(Region))

# Isolate Unmatched
unmatched <- joined %>%
  filter(is.na(Region))

# 2 of 2 - left_join with xref
bcm_xreference_alt <- bcm_xreference %>%
  rename(alt_pgm_dgr_enroll_tmpl = pgm_dgr_enroll_tmpl)

find_alt_unmatch <- unmatched %>%
  left_join(bcm_xreference_alt, by = "alt_pgm_dgr_enroll_tmpl")

alt_unmatch_clean_1 <- find_alt_unmatch %>%
  select(-ends_with(".x"))

alt_unmatch_clean_2 <- alt_unmatch_clean_1 %>%
  rename_with(~ gsub("\\.y$", "", .), ends_with(".y"))

# Rejoin with Alternate Pairing
bind_matched_alt <- bind_rows(matched, alt_unmatch_clean_2)
bind_matched_alt <- clean_names(bind_matched_alt)
bind_matched_alt <- bind_matched_alt %>%
  rename(stu_enroll_type = student_type_2)

bind_matched_alt <- bind_matched_alt %>%
  filter(!is.na(enrollment_type))

na_region <- bind_matched_alt %>%
  filter(is.na(enrollment_type))

# write_csv(x = bind_matched_alt, file = "output/bcm_export.csv")
# write_csv(x = na_region, file = "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/output/bcm_na_export.csv")

bcm_term_seq <- bind_matched_alt %>%
  mutate(term_int = sapply(term, term_code_to_num))

bcm_term_seq_ex <- bcm_term_seq %>%
  mutate(cohort_int = sapply(cohort, term_code_to_num))
bcm_term__cl <- bcm_term_seq_ex %>%
  select(-pgm_dgr_enroll_tmpl, -alt_pgm_dgr_enroll_tmpl)
bcm_term__sum <- bcm_term__cl %>%
  group_by(
    term,
    term_int,
    cohort_int,
    region,
    enrollment_type,
    attendance_type,
    stu_enroll_type,
    program
  ) %>%
  summarise(
    row_count = n(),
    total_tuition_charge = sum(tuition_charge, na.rm = TRUE),
    total_fa_institutional = sum(fa_institutional, na.rm = TRUE),
    total_comp_fee = sum(comp_fee, na.rm = TRUE),
    total_ntr = (total_tuition_charge - total_fa_institutional) + total_comp_fee
  )
write_csv(bcm_term__sum, "bcm_term__sum.csv")
```

```{r  bcm_retention_prep}

bcm_term__g <- bcm_term__cl %>%
  mutate(session = case_when(
    substr(cohort_int, 5, 6) == "2" ~ "Fall",
    substr(cohort_int, 5, 6) == "3" ~ "Spring",
    substr(cohort_int, 5, 6) == "1" ~ "Summer"
  ))





glimpse(bcm_term__g)

write_csv(x = bcm_term__g, file = "/Volumes/Enrollment_Marketing_Analytics/Cohort Based Projections/bcm_export.csv")
# write_sheet(data = bcm_term__sum, ss = "https://docs.google.com/spreadsheets/d/1AO7M-oRt4RFaKss_AHhgSYnsITkAuc1VTFpVJnhaY6Q/edit?gid=989055146#gid=989055146", sheet = "bcm_historicals")

part_time_bcm <- bcm_term__g %>%
  filter(attendance_type == "Part Time")

write_csv(x = part_time_bcm, file = "/Volumes/Enrollment_Marketing_Analytics/Cohort Based Projections/part_time_bcm_export.csv")


bcm_term__g_su <- bcm_term__g %>%
  filter(term == "2024SU") %>%
  distinct()

write_csv(bcm_term__g_su, "bcm_term__g_su.csv")
```

```{r  bcb_net_revenue}
# Spring Group
{
  bcb_2025_spring <-
    read_csv(
      "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/raw_bcb_osa/bcb_2025_spring.csv"
    )
  bcb_2024_spring <-
    read_csv(
      "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/raw_bcb_osa/bcb_2024_spring.csv"
    )
  bcb_2023_spring <-
    read_csv(
      "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/raw_bcb_osa/bcb_2023_spring.csv"
    )
  bcb_2022_spring <-
    read_csv(
      "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/raw_bcb_osa/bcb_2022_spring.csv"
    )
  bcb_2021_spring <-
    read_csv(
      "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/raw_bcb_osa/bcb_2021_spring.csv"
    )
  bcb_2020_spring <-
    read_csv(
      "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/raw_bcb_osa/bcb_2020_spring.csv"
    )
  bcb_2019_spring <-
    read_csv(
      "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/raw_bcb_osa/bcb_2019_spring.csv"
    )
  bcb_2018_spring <-
    read_csv(
      "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/raw_bcb_osa/bcb_2018_spring.csv"
    )
}

# Fall Group
{
  bcb_2024_fall <-
    read_csv(
      "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/raw_bcb_osa/bcb_2024_fall.csv"
    )
  bcb_2023_fall <-
    read_csv(
      "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/raw_bcb_osa/bcb_2023_fall.csv"
    )
  bcb_2022_fall <-
    read_csv(
      "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/raw_bcb_osa/bcb_2022_fall.csv"
    )
  bcb_2021_fall <-
    read_csv(
      "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/raw_bcb_osa/bcb_2021_fall.csv"
    )
  bcb_2020_fall <-
    read_csv(
      "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/raw_bcb_osa/bcb_2020_fall.csv"
    )
  bcb_2019_fall <-
    read_csv(
      "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/raw_bcb_osa/bcb_2019_fall.csv"
    )
  bcb_2018_fall <-
    read_csv(
      "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/raw_bcb_osa/bcb_2018_fall.csv"
    )
}
bcb_2020_fall$`Net Revenue` <-
  as.numeric(bcb_2020_fall$`Net Revenue`)
bcb_2025_spring$PEOPLE_ID <- as.numeric(bcb_2025_spring$PEOPLE_ID)


#
```

```{r  bcb_retention_prep}


# Bind Spring and Fall together
bcb_ntr_data_sp <-
  bind_rows(
    bcb_2025_spring,
    bcb_2024_spring,
    bcb_2023_spring,
    bcb_2022_spring,
    bcb_2021_spring,
    bcb_2020_spring,
    bcb_2019_spring,
    bcb_2018_spring,
    .id = "spring"
  ) %>%
  distinct()

bcb_ntr_data_fa <-
  bind_rows(
    bcb_2024_fall,
    bcb_2023_fall,
    bcb_2022_fall,
    bcb_2021_fall,
    bcb_2020_fall,
    bcb_2019_fall,
    bcb_2018_fall,
    .id = "fall"
  ) %>%
  distinct()

# xxx
bcb_ntr_data <-
  bind_rows(bcb_ntr_data_sp, bcb_ntr_data_fa) %>%
  distinct() %>%
  clean_names()



write_csv(
  bcb_ntr_data,
  "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/merged_bcb_osa/bcb_ntr_data.csv"
)


# STOP to check

# bcb_ntr_data <- read_csv(
#    "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/merged_bcb_osa/bcb_ntr_data.csv"
# )
str(bcb_ntr_data)
bcb_ntr_cl <- bcb_ntr_data %>%
  select(
    -spring,
    -entering_student,
    -applied_lesson_number,
    -applied_lesson_charge,
    -lens_course,
    -resident_commuter,
    -food_plan,
    -meal_charge,
    -dorm_plan,
    -housing_charge,
    -health_charge,
    -health_waiver,
    -part_time,
    -tuition_comp,
    -net_revenue,
    -ext_fed_aid,
    -single_room_charge,
    -fall
  ) %>%
  distinct()
# left_join with xref table
bcb_xreference <- read_csv(
  "/Volumes/Enrollment_Marketing_Analytics/Enrollment Models/Net_Tuition_Revenue/osa_data/bcb_osa/bcb_xreference.csv"
)
cols_bcb <- bcb_ntr_cl %>%
  mutate(pgm_dgr_enroll_tmpl = paste0(program, degree, enrollment_source)) %>%
  distinct()

bcb_xref_cl <- bcb_xreference %>%
  clean_names()

bcb_cohorts <- cols_bcb %>%
  left_join(bcb_xref_cl, by = "pgm_dgr_enroll_tmpl")

bcb_match <- bcb_cohorts %>%
  filter(!is.na(region_xref))

bcb_no_match <- bcb_cohorts %>%
  filter(is.na(region_xref)) %>%
  select(
    -pgm_dgr_enroll_tmpl,
    -region_xref,
    -enroll_type_xref,
    -attend_type_xref,
    -student_type_xref,
    -program_xref
  )

bcb_alts <- bcb_no_match %>%
  mutate(pgm_dgr_enroll_tmpl = paste0(program, degree, enrollment_source, curriculum)) %>%
  distinct()

bcb_alts_match <- bcb_alts %>%
  left_join(bcb_xref_cl, by = "pgm_dgr_enroll_tmpl")

# bcb_alts_ma_cl1 <- bcb_alts_match %>%
#    rename_with( ~ gsub("\\.y$", "", .), ends_with(".y"))

bcb_combine <- bind_rows(bcb_match, bcb_alts_match)

bcb_cl1 <- bcb_combine %>%
  mutate(
    admit_term = case_when(is.na(admit_term) ~ academic_term, TRUE ~ admit_term),
    admit_year = case_when(is.na(admit_year) ~ academic_year, TRUE ~ academic_year)
  )

table(bcb_cl1$admit_term, bcb_cl1$admit_year)
table(bcb_cl1$academic_term, bcb_cl1$academic_year)

# term_code conversion
bcb_cl2 <- bcb_cl1 %>%
  mutate(
    term_code = paste0(academic_year, substr(academic_term, 1, 2))
  )
table(bcb_cl2$term_code)

# Term Conversion / Consistency
admit_cohort <- bcb_cl2 %>%
  mutate(cohort = case_when(
    is.na(admit_year) ~ term_code,
    !is.na(admit_year) ~ paste0(admit_year, substr(admit_term, 1, 2))
  ))
table(admit_cohort$cohort)

admit_cohort_int <- admit_cohort %>%
  mutate(cohort_int = sapply(cohort, term_code_to_num))

table(admit_cohort_int$cohort_int)
write_csv(x = admit_cohort, file = "/Volumes/Enrollment_Marketing_Analytics/Cohort Based Projections/bcb_export.csv")

bcb_term__sum <- admit_cohort %>%
  group_by(
    region_xref,
    term_code,
    admit_term,
    admit_year,
    enroll_type_xref,
    attend_type_xref,
    student_type_xref,
    program_xref
  ) %>%
  summarise(
    row_count = n(),
    total_tuition_charge = sum(tuition_charge, na.rm = TRUE),
    total_fa_institutional = sum(scholarships, na.rm = TRUE),
    total_comp_fee = sum(comp_charge, na.rm = TRUE),
    total_ntr = (total_tuition_charge - total_fa_institutional) + total_comp_fee
  )
write_csv(bcb_term__sum, "bcb_term__sum.csv")

# write_sheet(
#   bcb_term__sum,
#   "https://docs.google.com/spreadsheets/d/1AO7M-oRt4RFaKss_AHhgSYnsITkAuc1VTFpVJnhaY6Q/edit?gid=749683695#gid=749683695",
#   "bcb_historicals"
# )
```
